## Gene length Analysis

This notebook processes the extracted data from `.txt` files and generates visualizations.

```{r}
# Load necessary libraries
library(tidyverse)
library(readxl)
library(dplyr)
library(conflicted)
conflict_prefer("filter", "dplyr")

```

## Set the organisms you want to process

```{r}
# Create a data frame for the UP numbers and organisms
up_to_organism <- tibble(
  UP_number = c("UP000000625", "UP000002311", "UP000006548", "UP000001940", "UP000000803", "UP000000539", "UP000002279", "UP000002280", "UP000001075", "UP000000589", "UP000006718", "UP000001595", "UP000001519", "UP000002277", "UP000005640"),
  Organism = c("Escherichia coli", "Saccharomyces cerevisiae", "Arabidopsis thaliana", "Caenorhabditis elegans", "Drosophila melanogaster", "Gallus gallus", "Ornithorhynchus anatinus", "Monodelphis domestica", "Cricetulus griseus", "Mus musculus", "Macaca mulatta", "Pongo abelii", "Gorilla gorilla gorilla", "Pan troglodytes", "Homo sapiens")
)

# Display the table using knitr::kable
knitr::kable(up_to_organism, caption = "Mapping of UP Numbers to Organisms")

```

## Select organisms you want to display here.

All organisms with their UP IDs in this list will be processed and shown in the figures.

```{r}
# Select only the organisms you want to plot
selected_organisms <- c("UP000000625", "UP000002311", "UP000006548", "UP000001940", "UP000000803", "UP000000539", "UP000002279", "UP000002280", "UP000001075", "UP000000589", "UP000006718", "UP000001595", "UP000001519", "UP000002277", "UP000005640")
```

## Load Data

Reading the `.tsv` files into R and combining them into a single dataframe.

```{r}
# Read in data from .tsv files
file_paths <- list.files("D:/Emilia/low_complexity_project/analysis/genelengthoutput", 
                         pattern = "*.tsv", full.names = TRUE)
organism_data <- file_paths %>%
  map_dfr(~read_tsv(.x, show_col_types = FALSE) %>%
            mutate(Organism = str_remove(tools::file_path_sans_ext(basename(.x)), "_length")))

HkPolyx <- organism_data %>%
  filter(Organism %in% selected_organisms) %>%  # Filter data for selected organisms
  droplevels() %>%
  filter(genelength != "-") %>%
  mutate(Organism = factor(Organism)) %>%
  mutate(Hk = recode(Hk, `0` = "not Hk", `1` = "Hk"),
         Hk = as.factor(Hk),
         Hk = factor(Hk, levels = rev(levels(Hk))),
         genelength = as.numeric(genelength)) %>%
  filter(!is.na(genelength))

print(unique(HkPolyx$Organism))

```
## Link UP IDs to scientific organism names

```{r}
# organism names
organism_labels <- c(
  "UP000000625" = "italic('E. coli')",
  "UP000002311" = "italic('S. cerevisiae')",
  "UP000001940" = "italic('C. elegans')",
  "UP000000803" = "italic('D. melanogaster')",
  "UP000000589" = "italic('M. musculus')",
  "UP000005640" = "italic('Homo sapiens')",
  "UP000006548" = "italic('A. thaliana')",
  "UP000006718" = "italic('Macaca mulatta (28.8 MYA)')",
  "UP000001595" = "italic('Pongo abelii (15.2 MYA)')",
  "UP000001519" = "italic('Gorilla gorilla gorilla (8.6 MYA)')",
  "UP000002277" = "italic('Pan troglodytes (6.4 MYA)')",
  "UP000000539" = "italic('Gallus gallus (319 MYA)')",
  "UP000002279" = "italic('O. anatinus')",      # (180 MYA)
  "UP000002280" = "italic('M. domestica')",     #  (160 MYA)
  "UP000001075" = "italic('C. griseus (87 MYA)')"
)
```

## Print average gene length

```{r}
# Convert Organism column AFTER filtering
HkPolyx$Organism <- factor(HkPolyx$Organism, levels = selected_organisms, labels = organism_labels[selected_organisms])
HkPolyx

# Group and calculate average gene length
average_lengths <- HkPolyx %>%
  group_by(Organism, Hk) %>%
  summarize(average_genelength = mean(genelength, na.rm = TRUE),
            .groups = "drop")

# Print the result
print(average_lengths)


# Summarize counts and mean lengths per Hk status
by_status <- HkPolyx %>%
  group_by(Organism, Hk) %>%
  summarize(
    Count = n(),
    Mean_Length = mean(genelength, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = Hk,
    values_from = c(Count, Mean_Length),
    names_glue = "{.value}_{Hk}"
  )

# Add overall mean gene length per organism (ignoring Hk status)
overall_means <- HkPolyx %>%
  group_by(Organism) %>%
  summarize(Mean_Length_All = mean(genelength, na.rm = TRUE), .groups = "drop")

# Join both tables
final_summary <- left_join(by_status, overall_means, by = "Organism")

# Show as a nice table
knitr::kable(final_summary, digits = 2, caption = "Gene counts and average lengths by organism and Hk status")


```

## Assign significance values manually

```{r}
# manual significance values
significance_values <- c(
  "UP000000625" = "***",
  "UP000006548" = "***",
  "UP000002311" = "***",
  "UP000001940" = "***",
  "UP000000803" = "***",
  "UP000000589" = "***",
  "UP000005640" = "***",
  "UP000006718" = "***",
  "UP000001595" = "***",
  "UP000001519" = "**",
  "UP000002277" = "***",
  "UP000000539" = "ns",
  "UP000002279" = "***",
  "UP000002280" = "***",
  "UP000001075" = "***"
)


significance_filtered <- tibble(
  Organism = factor(
    selected_organisms, 
    levels = selected_organisms, 
    labels = organism_labels[selected_organisms]
  ),
  x = rep(1.5, length(selected_organisms)),
  y = rep(80000, length(selected_organisms)), 
  label = significance_values[selected_organisms]
)
```

## Visualization
```{r}
# Boxplot visualization with human-readable names
HkPolyx %>%
  ggplot(aes(x = Hk, y = genelength, fill = Hk)) +
  geom_boxplot(size = 1.2, notch = TRUE) +
  ylab("Gene length (bp)") +
  coord_cartesian(ylim = c(0, 5000)) +                   # Adjust limits depending on organisms gene length.
  scale_fill_manual(values = c("#ffc8d2", "#ccffee")) +
  stat_summary(fun = "mean", geom = "point", shape = 21, size = 4, color = "black", fill = "white") +
  facet_wrap(~Organism, ncol = 2, labeller = label_parsed) +
  theme_minimal() +
  geom_text(
    data = HkPolyx %>%
      group_by(Hk, Organism) %>%
      summarize(median_length = median(genelength, na.rm = TRUE),
                Count = n(), .groups = "drop"),
    aes(x = Hk,
        y = median_length - 5000,  # Adjust position as needed
        label = Count),
    inherit.aes = FALSE,
    size = 3, color = "black") +
  geom_text(data = significance_filtered, aes(x = x, y = y, label = label), size = 5, color = "black", inherit.aes = FALSE)

# E coli, yeast and arabidopsis: very short genes (< 3000)
# C elegans and D melanogaster: short genes (< 7000)
# All mammals: Long genes (40.000 - 80.000)

```

## Test for significance

```{r}
# Wilcoxon Rank-Sum Test for significance
plot1_stats <- HkPolyx %>%
  group_by(Organism) %>%
  summarise(
    p_value = wilcox.test(genelength ~ Hk)$p.value
  ) %>%
  mutate(significance = case_when(
    p_value < 0.001 ~ "***",
    p_value < 0.01 ~ "**",
    p_value < 0.05 ~ "*",
    TRUE ~ "ns"
  ))

print(plot1_stats)
```

