## HR, HK, and PolyX Analysis

This notebook processes the extracted data from `.txt` files and generates visualizations.

```{r}
# Load necessary libraries
library(tidyverse)
library(readxl)
library(dplyr)
library(conflicted)
conflict_prefer("filter", "dplyr")

```

## Set the organisms you want to process

```{r}
# Create a data frame for the UP numbers and organisms
up_to_organism <- tibble(
  UP_number = c("UP000000625", "UP000002311", "UP000006548", "UP000001940", "UP000000803", "UP000000539", "UP000002279", "UP000002280", "UP000001075", "UP000000589", "UP000006718", "UP000001595", "UP000001519", "UP000002277", "UP000005640"),
  Organism = c("Escherichia coli", "Saccharomyces cerevisiae", "Arabidopsis thaliana", "Caenorhabditis elegans", "Drosophila melanogaster", "Gallus gallus", "Ornithorhynchus anatinus", "Monodelphis domestica", "Cricetulus griseus", "Mus musculus", "Macaca mulatta", "Pongo abelii", "Gorilla gorilla gorilla", "Pan troglodytes", "Homo sapiens")
)

# Display the table using knitr::kable
knitr::kable(up_to_organism, caption = "Mapping of UP Numbers to Organisms")

```

## Select organisms you want to display here.

All organisms with their UP IDs in this list will be processed and shown in the figures.

```{r}
# Select only the organisms you want to plot
selected_organisms <- c("UP000000625", "UP000002311", "UP000006548", "UP000001940", "UP000000803", "UP000000539", "UP000002279", "UP000002280", "UP000001075", "UP000000589", "UP000006718", "UP000001595", "UP000001519", "UP000002277", "UP000005640")

```

## Load Data

Reading the `.tsv` files into R and combining them into a single dataframe.

```{r}
# Read in data from .tsv files
file_paths <- list.files("./proteomes_hrs_hk", pattern = "*.tsv", full.names = TRUE)
organism_data <- file_paths %>%
  map_dfr(~read_tsv(.x, show_col_types = FALSE) %>%
            mutate(
              Polyx_lengths = as.character(Polyx_lengths),
              Count_grouped = as.character(Count_grouped),
              Organism = str_remove(tools::file_path_sans_ext(basename(.x)), "_hrs_hk")
            ))

HKPolyx <- organism_data %>%
  filter(Organism %in% selected_organisms) %>%  # Filter data for selected organisms
  droplevels() %>%
  mutate(Organism = factor(Organism))

head(HKPolyx)

print(unique(HKPolyx$Organism))

```

## Data Processing

Processing the data to classify HK and count groups properly.

```{r}
HKPolyx <- HKPolyx %>%
  mutate(HK = recode(HK, `0` = "not HK", `1` = "HK"),
         HK = as.factor(HK),
         HK = factor(HK, levels = rev(levels(HK))),
         Count_grouped = factor(Count_grouped, levels = c("0", "1", ">1")))


# Table with all numbers
by_status <- HKPolyx %>%
  group_by(Organism, HK) %>%
  summarize(
    Count = n(),
    Mean_Length = mean(Length, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = HK,
    values_from = c(Count, Mean_Length),
    names_glue = "{.value}_{HK}"
  )

overall_means <- HKPolyx %>%
  group_by(Organism) %>%
  summarize(Mean_Length_All = mean(Length, na.rm = TRUE), .groups = "drop")
final_summary <- left_join(by_status, overall_means, by = "Organism")

knitr::kable(final_summary, digits = 2, caption = "Protein counts and average lengths by organism and HK status")
```

## Mutate column "Count_grouped" to either 0, 1 or \>1 homorepeats

```{r}
HKPolyx <- HKPolyx %>%
  mutate(
    Count_grouped = case_when(
      Polyx_count == "0" ~ "0",
      Polyx_count == "1" ~ "1",
      as.numeric(Polyx_count) > 1 ~ ">1",  # for values greater than 2
      TRUE ~ NA_character_  # handle any unexpected values
    ),
    Count_grouped = factor(Count_grouped, levels = c("0", "1", ">1"))
  )
```

## Link UP IDs to scientific organism names

```{r}
# organism names
organism_labels <- c(
  "UP000000625" = "italic('E. coli')",
  "UP000002311" = "italic('S. cerevisiae')",
  "UP000001940" = "italic('C. elegans')",
  "UP000000803" = "italic('D. melanogaster')",
  "UP000000589" = "italic('M. musculus')",
  "UP000005640" = "italic('Homo sapiens')",
  "UP000006548" = "italic('A. thaliana')",
  "UP000006718" = "italic('Macaca mulatta (28.8 MYA)')",
  "UP000001595" = "italic('Pongo abelii (15.2 MYA)')",
  "UP000001519" = "italic('Gorilla gorilla gorilla (8.6 MYA)')",
  "UP000002277" = "italic('Pan troglodytes (6.4 MYA)')",
  "UP000000539" = "italic('Gallus gallus (319 MYA)')",
  "UP000002279" = "italic('O. anatinus')",      # (180 MYA)
  "UP000002280" = "italic('M. domestica')",     #  (160 MYA)
  "UP000001075" = "italic('C. griseus (87 MYA)')"
)
```

## HK Visualization - Protein length by HK

```{r}

# Convert Organism column AFTER filtering
HKPolyx$Organism <- factor(HKPolyx$Organism, levels = selected_organisms, labels = organism_labels[selected_organisms])

# manual significance values
significance_values_HK <- c(
  "UP000000625" = "***",
  "UP000002311" = "***",
  "UP000001940" = "***",
  "UP000000803" = "***",
  "UP000000589" = "***",
  "UP000006548" = "***",
  "UP000005640" = "***",
  "UP000006718" = "***",
  "UP000001595" = "*",
  "UP000001519" = "ns",
  "UP000002277" = "ns",
  "UP000000539" = "ns",
  "UP000002279" = "***",
  "UP000002280" = "***",
  "UP000001075" = "***"
)


significance_HK_filtered <- tibble(
  Organism = factor(
    selected_organisms, 
    levels = selected_organisms, 
    labels = organism_labels[selected_organisms]
  ),
  x = rep(1.5, length(selected_organisms)),
  y = rep(850, length(selected_organisms)),
  label = significance_values_HK[selected_organisms]
)

# final plot
HKPolyx %>%
  ggplot(aes(x = HK, y = Length, fill = HK)) +
  geom_boxplot(size = 1.2, notch = TRUE) +
  ylab("Protein length (aa)") +
  coord_cartesian(ylim = c(150, 850)) +
  scale_fill_manual(values = c("#ffc8d2", "#ccffee")) +
  stat_summary(fun = "mean", geom = "point", shape = 21, size = 4, color = "black", fill = "white") +
  facet_wrap(~Organism, ncol = 4, labeller = label_parsed, drop = TRUE) +
  theme_minimal() +
  geom_text(
    data = HKPolyx %>%
      group_by(HK, Organism) %>%
      summarize(median_length = median(Length),
                Count = n()),
    aes(x = HK,
        y = median_length - 50,  # Place the text at the median
        label = Count),
    inherit.aes = FALSE,
    size = 3, color = "black") +
  geom_text(data = significance_HK_filtered, aes(x = x, y = y, label = label), size = 5, color = "black", inherit.aes = FALSE)

```

## HR Visualization - Protein Length by HR

Generating a boxplot of protein length grouped by the number of homorepeats.

```{r}

# Manual significance levels
significance_0_1 <- c(
  "UP000000625" = "ns",
  "UP000002311" = "***",
  "UP000001940" = "***",
  "UP000000803" = "***",
  "UP000000589" = "***",
  "UP000006548" = "***",
  "UP000005640" = "***",
  "UP000002277" = "***",
  "UP000006718" = "***",
  "UP000001519" = "***",
  "UP000001595" = "***",
  "UP000000539" = "***",
  "UP000002279" = "***",
  "UP000002280" = "***",
  "UP000001075" = "***"
)

significance_1_2 <- c(
  "UP000002311" = "***",
  "UP000001940" = "***",
  "UP000000803" = "***",
  "UP000000589" = "***",
  "UP000006548" = "***",
  "UP000005640" = "***",
  "UP000002277" = "***",
  "UP000006718" = "***",
  "UP000001519" = "***",
  "UP000001595" = "***",
  "UP000000539" = "***",
  "UP000002279" = "***",
  "UP000002280" = "***",
  "UP000001075" = "ns"
)


# Create the significance label table
significance_HR_filtered <- tibble()

for (org in selected_organisms) {
  org_label <- factor(org, levels = selected_organisms, labels = organism_labels[selected_organisms])
  
  # 0 vs 1 (include all)
  significance_HR_filtered <- bind_rows(significance_HR_filtered,
    tibble(
      Organism = org_label,
      x = 1.5,
      y = 1500,
      label = significance_0_1[org]
    )
  )
  
  # 1 vs 2 (exclude E. coli)
  if (!is.null(significance_1_2[org])) {
    significance_HR_filtered <- bind_rows(significance_HR_filtered,
      tibble(
        Organism = org_label,
        x = 2.5,
        y = 1500,
        label = significance_1_2[org]
      )
    )
  }
}

# Final plot
HKPolyx %>%
  ggplot(aes(x = Count_grouped, y = Length, fill = Count_grouped)) +
  geom_boxplot(size = 1.2, notch = TRUE) +
  xlab("Number of Homorepeats") +
  ylab("Protein length (aa)") +
  coord_cartesian(ylim = c(100, 1600)) +
  scale_fill_manual(name = "HR count", values = c("#4d79ff", "#ccffee", "#ffffff")) +
  stat_summary(fun = "mean", geom = "point", shape = 21, size = 4, color = "black", fill = "white") +
  facet_wrap(~Organism, ncol = 3, labeller = label_parsed, drop = TRUE) +
  theme_minimal() +
  geom_text(data = significance_HR_filtered,
            aes(x = x, y = y, label = label),
            size = 4, color = "black", inherit.aes = FALSE)
```

## Check p-values for significance with Wilcoxon test

```{r}
# Wilcoxon Rank-Sum Test for Plot 1: Protein Length by Housekeeping Status
plot1_stats <- HKPolyx %>%
  group_by(Organism) %>%
  summarise(
    p_value = wilcox.test(Length ~ HK)$p.value
  ) %>%
  mutate(significance = case_when(
    p_value < 0.001 ~ "***",
    p_value < 0.01 ~ "**",
    p_value < 0.05 ~ "*",
    TRUE ~ "ns"
  ))

print(plot1_stats)

# Apply Wilcoxon test for each organism for 0 vs. 1
plot2_0_1 <- HKPolyx %>%
  group_by(Organism) %>%
  filter(Count_grouped %in% c("0", "1")) %>%
  do({
    # Perform the Wilcoxon test for the current organism
    test_result <- wilcox.test(Length ~ Count_grouped, data = ., correct = TRUE)
    # Return the p-value for the test
    tibble(p_value_0_1 = test_result$p.value)
  })

# View the results
print(plot2_0_1)

# Apply Wilcoxon test for each organism for 1 vs. >1
plot2_1_2 <- HKPolyx %>%
  filter(Organism != "italic('E. coli')") %>%
  group_by(Organism) %>%
  filter(Count_grouped %in% c("1", ">1")) %>%
  do({
    # Perform the Wilcoxon test for the current organism
    test_result <- wilcox.test(Length ~ Count_grouped, data = ., correct = TRUE)
    # Return the p-value for the test
    tibble(p_value_1_2 = test_result$p.value)
  })

print(plot2_1_2)
```
## Number of HRs in HK and non-HK proteins for each organism

```{r}
expanded_HKPolyx <- HKPolyx %>%
  group_by(HK, Organism) %>%
  separate_rows(Polyx_types, Polyx_lengths, sep = "/") %>%
  filter(Polyx_types != "-") %>%
  summarise(Count = n(), .groups = "drop")
expanded_HKPolyx
```
# Number of HR-containing proteins

```{r}
HKPolyx
PHRs <- HKPolyx %>%
  filter(Polyx_count > 0) %>%                            # Only proteins with at least one polyX
  group_by(Organism, HK) %>%
  summarise(Proteins_with_polyX = n(), .groups = "drop") %>%
  pivot_wider(names_from = HK, values_from = Proteins_with_polyX, values_fill = 0) %>%
  rename(
    HK_with_polyX = `HK`,
    non_HK_with_polyX = `not HK`
  )
PHRs
```

